cardinalR: Generating Interesting High-Dimensional Data Structures

### 2025-12-11 paper-cardinalR.Rmd
Added sample for model-based clusters

```{r fig-cluster-stats, fig.width=10, fig.height=3, out.width="100%", layout="l-body"}
# Examine the cluster stats
all_stats <- bind_rows(hc_stats, kmeans_stats)
all_stats |>
  pivot_longer(within.cluster.ss:sindex, names_to = "stat", values_to = "value") |>
  ggplot(aes(x=cl, y=value, colour=method)) +
    geom_line() +
    facet_wrap(~stat, ncol=3, scales="free_y") +
    xlab("Number of clusters") +
    ylab("")
```

@fig-cluster-stats shows a selection of cluster metrics for 2-10 clusters for each of the methods, $k$-means, hierarchical, ... As is typical, the suggestion of best solution varies between cluster statistics. There is some agreement that 4-5 clusters is better than other choices (say why), and that $k$-means performs better than others (this might not hold when you add the next methods).

>>>>

```{r}
# Model-based clustering
modelbased_cl <- NULL
for (ncl in 2:10) {
  mcl <- Mclust(five_clusts[, -5], G = ncl, modelNames = "VVV")$classification
  modelbased_cl <- cbind(modelbased_cl, mcl)
}
colnames(modelbased_cl) <- paste0("cl", 2:10)

x <- cluster.stats(dist_mat,
  modelbased_cl[,1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
modelbased_stats <- tibble(method="mb", cl=2) |>
  bind_cols(as_tibble(x))
for (i in 3:10) {
  x <- cluster.stats(dist_mat,
    modelbased_cl[,i-1])[c("within.cluster.ss", "pearsongamma", "dunn2", 
                    "wb.ratio", "ch", "sindex")]
  mbs <- tibble(method="mb", cl=i) |>
    bind_cols(as_tibble(x))
  modelbased_stats <- modelbased_stats |>
    bind_rows(mbs)
}
```

```{r, label = "fig-cluster-stats", fig.width=10, fig.height=3, out.width="100%", layout="l-body"}
# Examine the cluster stats
all_stats <- bind_rows(hc_stats, kmeans_stats, modelbased_stats)
all_stats |>
  pivot_longer(within.cluster.ss:sindex, names_to = "stat", values_to = "value") |>
  ggplot(aes(x=cl, y=value, colour=method)) +
    geom_line() +
    facet_wrap(~stat, ncol=3, scales="free_y") +
    xlab("Number of clusters") +
    ylab("") +
    theme_minimal() +
    theme(panel.border = element_rect(fill = 'transparent'),
          plot.title = element_text(size = 12, hjust = 0.5, vjust = -0.5),
          axis.ticks.x = element_line(),
          axis.ticks.y = element_line(),
          legend.position = "none",
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          axis.title.x = element_text(size = 7),
          axis.title.y = element_text(size = 7),
          plot.margin = margin(0, 0, 0, 0))
```

@fig-cluster-stats shows a selection of cluster metrics for $2-10$ clusters for each of the methods, $k$-means, hierarchical, model-based. As is typical, the suggestion of best solution varies between cluster statistics. There is some agreement that $4-5$ clusters is better than other choices (say why), and that $k$-means performs better than others (this might not hold when you add the next methods).

### 2025-12-11 paper-cardinalR.tex
Rendered new version
Added sample for model-based clusters

\subsection{Generating high-dimensional clustered data}\label{generating-high-dimensional-clustered-data}

@@

\citet{fig-cluster-stats} shows a selection of cluster metrics for 2-10 clusters for each of the methods, \(k\)-means, hierarchical, \ldots{} As is typical, the suggestion of best solution varies between cluster statistics. There is some agreement that 4-5 clusters is better than other choices (say why), and that \(k\)-means performs better than others (this might not hold when you add the next methods).

>>>>

@@

\citet{fig-cluster-stats} shows a selection of cluster metrics for \(2-10\) clusters for each of the methods, \(k\)-means, hierarchical, model-based. As is typical, the suggestion of best solution varies between cluster statistics. There is some agreement that \(4-5\) clusters is better than other choices (say why), and that \(k\)-means performs better than others (this might not hold when you add the next methods).
